The Anatomy of a Hit

### 2024-04-25 paper/paper.qmd
edited paper (for the last time, hopefully)

---
title: "The Anatomy of a Hit"
subtitle: "Modeling Popularity of Billboard Artists using Spotify Audio Features"
author: 
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/popularity-modeling. Thank you to Michaela Drouillard for your insights on Spotify and it's metrics. My sincere appreciation goes to Hannah Yu for your valuable last-minute edits. Special hat tip goes to Rajan Maghera and my amazing parents, Cristina and Shane for your support and encouragement. I could not have done this without you."
date: April 24 2024
date-format: long
abstract: "What makes a hit pop song? This paper examines the audio features that characterize mainstream music's biggest hits. By analyzing data from the discographies of the greatest artists on the Billboard Hot 100, the research identifies key attributes associated with popular songs. Multivariate regression analysis reveals that higher levels of danceability, explicit lyrics, and loudness are positively related to popularity, while emotional positivity (valence) exhibits a negative relationship. Overall, this work quantifies some factors that drive iconic pop successes and can empower professional pop musicians to make more informed creative decisions with their work."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
nocite: |
  @*
---

@@

Taken together, the regression results indicate that higher Spotify popularity scores are associated with danceability, explicit lyrics, loudness, and to a lesser extent, song duration, while valence (emotional positivity) has a negative association. The musical mode does not appear to have a significant effect on popularity when other predictors are accounted for.

@@

Armed with the knowledge of what elements work in making music popular, artists and music creators can be more empowered in their creative process. Using the insights from my analysis, they can be enabled to strategically focus their creative efforts and artistic vision on incorporating these elements, potentially mitigating common challenges such as writer's block or creative stagnation. They could also choose to diverge from these elements in order to explore new genres or music styles that may not be popular now, but could be in the future. Knowing the current trends now could inspire artists to make better art that effectively and innovatively builds on what has come before.

@@

Additionally, the crowdsourced nature of certain audio features like valence and danceability, initially derived from subjective assessments by college interns @drouillard2024, introduces a clear element of human subjectivity and potential bias in the collection of data, which could undermine the validity and generalizability of the relationships observed between these features and popularity scores. The reliance solely on Spotify popularity scores as the dependent variable could bias the results toward newer music versus older, more established releases, although the recency factor incorporated into Spotify's scoring algorithm could potentially mitigate this concern for tracks that have maintained enduring popularity over time. In particular, given the little amount of open and reliable information about how the Spotify popularity score and other generated metrics are calculated, it is worth questioning whether if those scores are associated with popularity scores by design.

There could also be potential methodological issues stemming from the complex and interrelated nature of the data itself. Most obviously, potential multicollinearity among the predictor variables (audio features) could affect the interpretation and relative influence of individual variables on popularity scores.

Moreover, the use of a linear regression model, while providing initial insights, could oversimplify the relationships by failing to capture non-linear or interaction effects among the variables. The dataset's limited scope, comprising only the top 10 songs per artist and excluding some tracks due to missing audio feature data from Spotify's API, restricts the sample size and may impact the precision and robustness of the parameter estimates.

@@

To address the weaknesses and expand upon the present findings, there are several avenues for future research emerge.

@@

Repeating this analysis on specific musical genres, focusing on particular artists, or using different definitions of success in popular music could also uncover genre-specific relationships or other variable influences that may be still obscured despite in light of me doing this work.

My approach affords merit as the first step by a cultural analyst in using the openly available data from Spotify to analyze and quantify the elements of what could make popular music popular. Through this analysis, cultural analysts should be inspired to continue work uncovering the elements that make up popular pop music. Artists also should be inspired to create better, more interesting art, rather than be afraid of their work being reduced down to a number; doing statistics is about simplifying, rather than a completely describing the world. Although much was learned in this analysis, there is still much to discover about the uniquely human elements of what makes our global culture tick.

>>>>

---
title: "The Anatomy of a Hit"
subtitle: "Modeling Popularity of Billboard Artists using Spotify Audio Features"
author: 
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/popularity-modeling. Thank you to Michaela Drouillard for your insights on Spotify and it's metrics. My sincere appreciation goes to Hannah Yu for your valuable last-minute edits. Special hat tips go to Rajan Maghera and my amazing parents, Cristina and Shane for your support and encouragement. I could not have done this without you all of your support."
date: April 24 2024
date-format: long
abstract: "What makes a hit pop song? This paper examines the audio features that characterize mainstream music's biggest hits. By analyzing data from the discographies of the greatest artists on the Billboard Hot 100, the research identifies key attributes associated with popular songs. Multivariate regression analysis reveals that higher levels of danceability, explicit lyrics, and loudness are positively related to popularity, while emotional positivity (valence) exhibits a negative relationship. Overall, this work quantifies some factors that drive iconic pop successes and can empower professional pop musicians to make more informed creative decisions with their work."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
nocite: |
  @*
---

@@

The regression results indicate that higher Spotify popularity scores are associated with danceability, explicit lyrics, loudness, and to a lesser extent, song duration, while valence (emotional positivity) has a negative association. On the other hand, musical mode does not appear to have a significant effect on popularity when other predictors are accounted for.

@@

Armed with the knowledge of what elements work in making music popular, artists and music creators can be more empowered in their creative process. Using the insights from my analysis, they can be enabled to strategically focus their creative efforts and artistic vision on incorporating these elements, potentially mitigating common challenges such as writer's block or creative stagnation. They could also choose to diverge from these elements in order to explore new genres or music styles that may not be popular now, but could be in the future. Knowing the current trends now could inspire artists to make better art that effectively innovates on what has come before.

@@

Additionally, the crowdsourced nature of certain audio features like valence and danceability, initially derived from subjective assessments by college interns @drouillard2024, introduces a clear element of human subjectivity and potential bias in the collection of data, which could undermine the validity and generalizability of the relationships observed between these features and popularity scores. The reliance solely on Spotify popularity scores as the dependent variable could bias the results toward newer music versus older, more established releases; Spotify is a digital-first company, meaning that their records do not include performance of songs that were released before their audio feature API went live in the early 2010s. That said, the recency factor incorporated into Spotify's scoring algorithm for popularity could potentially mitigate this concern for tracks that have maintained enduring popularity over time. Given the little amount of open and reliable information about how the Spotify popularity score and other generated metrics are calculated, there is much to question about the creation of these metrics, even considering whether these metrics were meant to be associated by design.

There are possible methodological issues in this design, primarily stemming from the complex and interrelated nature of the data itself. Most obviously, potential multicollinearity among the predictor variables (audio features) could affect the interpretation and relative influence of individual variables on popularity scores.  The use of a linear regression model, while providing initial insights, could have oversimplified relationships by failing to capture non-linear or interaction effects among the predictor variables. Given the dataset's limited scope, comprising only the top 10 songs per artist and excluding some tracks due to missing audio feature data from Spotify's API, means that the precision and robustness of the parameter estimates could likely be inaccurate and not generalizable.

@@

To address the weaknesses and expand upon the present findings, there are many, many emerging avenues for future research in this area.

@@

Finally, Repeating this analysis on specific musical genres, focusing on particular artists, or using different definitions of success in popular music could also uncover genre-specific relationships or other variable influences that may be still obscured despite in light of me doing this work.

My approach affords merit as the first step by a cultural analyst in using the openly available data from Spotify to analyze and quantify the elements of what could make popular music popular. Through this analysis, cultural analysts should be inspired to continue work uncovering the elements that make up popular pop music. Artists also should be inspired to create better, more interesting art, rather than be afraid of their work being reduced down to a number; doing statistics is about simplifying, rather than a completely describing the world. Although much was learned in this analysis, there is still much to discover about the uniquely human elements of what makes our globally connected culture tick.
